SmartCheck: Automatic and Efficient Counterexample Reduction and Generalization

### 2014-05-10 paper/paper.tex
paper: updates.

% can be 9pt
\documentclass{sigplanconf}
\usepackage{url}
\usepackage{tikz}
\usepackage{graphicx}
\usepackage{framed}
\usepackage{color}
\usepackage{url}
\usepackage{alltt}
\usepackage{tabularx}
\usepackage{subfigure}
\usepackage{fancyhdr}
\usepackage{paralist}
\usepackage{multirow}
\usepackage{pgf}

@@

\newcommand{\sci}[2]{\ensuremath{#1 \times 10^{#2}}}
\begin{document}

@@

That is, automated testing helps find a counterexample, but it may be a very
large counterexample.  We are left facing the \emph{shrinking problem}: assume
there exist a set of inputs $i_0, \ i_1, \ldots, \i_n$, such that for program
$P$ (modeled as a function mapping inputs to output) and property $\phi$,
$\phi(P(i_0, \ i_1, \ldots, \i_n))$ is false.  The shrinking problem is to find
the smallest set of inputs $i'_0, \ i'_1, \ldots, \i'_n$, according to some
measure, such that $\phi(P(i\_0, \ i\_1, \ldots, \i\_n))$ is also false.

@@

The third algorithm automatically strengthens properties by omitting ``similar''
counterexamples to the ones previously observed.  The algorithm is motivated by
noting that there are often multiple ways in which a property may fail; for
example, a property stating that pretty-printing an AST and then parsing it
results in the original AST may fail due to multiple bugs.  During testing, it
is useful to discover counterexamples arising from each bug in one go.  In
practice, the problem is solved by discovering a counterexample \ttp{cex},
abstracting it, and then adding a new precondition to the property that
informally says ``omit counterexamples of form \ttp{cex}.''  Adding
preconditions manually is laborious and may cause the programmer to make
premature fixes to the program, if she believes she has isolated the error
before she actually does.

@@

@@

\begin{table}[ht]
\footnotesize
  \begin{center}
    \begin{tabular}{|r||c|c|c|c|}
\hline
           & QC none     & QC trunc     & QC tuple    & SmartCheck\\
\hline \hline
Mean       & 83, 0.013s  & 40, 0.064s   & 10, 0.031s  & 6, 0.021s\\
\hline
Std. dev.  & 22, 0.014   & 20, 0.334    & 4,  0.039   & 4, 0.009\\
% Median     & 0.009s, 67  & 0.17s, 33  & 1.81s, 13  & 0.10s, 6\\
%% \hline
%% MAD        & 0.02s, 13  & 0.09s, 8   & 0.62s, 3   & 0.02s, 2\\
\hline
95\%       & 119, 0.023s  & 84, 0.153s  & 16, 0.052s  & 13, 0.036s\\
\hline
    \end{tabular}
  \end{center}
  \caption{Summarizing data for the graph in Figure~\ref{fig:graphs}. Entries
    contain execution time (in seconds) and counterexample sizes (counting
    \ttp{Int16} values).}
  \label{table:results}
\end{table}

The results of our discussion above are summarized in Table~\ref{table:results}
and the corresponding graph in Figure~\ref{fig:graphs}. They show three
approaches to shrinking with QuickCheck: no shrinking (QC none), to provide a
baseline, truncated shrinking (QC trunc) with \ttp{Int16} shrinking disabled, in
which each list is truncated to 10 elements, and tuple shrinking (QC tuple),
also with \ttp{Int16} shrinking disabled.  We omit SmallCheck, since as noted,
it is not even feasible for the example with default instances.\footnote{All
  results reported in the paper are with a GHC-7.6.3-compiled program, using
  \ttp{-O2}, on a 2.8 GHz 4~core i7 with 16GB memory under OSX v. 10.9.2.}  We
graph the final counterexample size, measured in the total number of \ttp{Int16}
values, and execution time for counterexample discovery and shrinking.
Figure~\ref{fig:graphs} shows how many of the 1000 runs resulted in a final
value of a particular size, fitting the data with a Bezier curve.

@@

class Arbitrary a => SubTypes a where
  size    :: a -> Size
  index   :: a -> Idx -> Maybe SubVal
  replace :: a -> Idx -> SubVal -> a
  constr  :: a -> String
  constrs :: a -> [String]
  opaque  :: a -> Bool
  subVals :: a -> Tree SubVal
\end{code}
%
\noindent
The \ttp{SubTypes} type class requires QuickCheck's \ttp{Arbitrary} as a
super-class.  \ttp{SubTypes} has the following methods:
\begin{itemize}
\item \ttp{size} returns the size of a value---intuitively, the number of
  constructors contained within it,
\item \ttp{index} returns a \emph{sub-value} at a breadth-first index in a
  value,
\item \ttp{replace} replaces a sub-value at a particular focus (returning the
  original value if the index is out-of-bounds),
\item \ttp{constr} returns a string representation of the constructor tagging
  the value,
\item \ttp{constrs} returns the list of all possible constructors from the
  value's type,
\item \ttp{opaque} is false when the type of the value is an ``interesting
  type''; informally, this is a type other than a primitive type like \ttp{Int},
  \ttp{Char}, or \ttp{Bool}.  See Section~\ref{sec:base} for a full discussion.
\item \ttp{subVals} returns a tree of all non opaque-type sub-values.  A tree is a
  value with the type
%
  \begin{code}
data Tree a = Node \{ rootLabel :: a
                   , subForest :: [Tree a] \}
  \end{code}
%
\end{itemize}

@@

subVals (B (B L\sub{0} L\sub{1}) L\sub{2}) =
  Node (B L\sub{0} L\sub{1}) [Node L\sub{0} [], Node L\sub{1} []]
       (Node L\sub{1} [])
\end{code}
%
\noindent
The \ttp{SubVal} type is an existential data type, used as a generic container
for sub-values from a counterexample.  We will sometimes refer to the unwrapped
value returned by \ttp{index a i} as the \emph{$i$th sub-value of $a$}, so for
example, (\ttp{B\sub{4} L\sub{6} L\sub{8}}) is the 4th sub-value of \ttp{tree}.
An invariant of \ttp{index} is that for any value \ttp{a}, and for the smallest
$i \geq 0$ such that
%
\begin{code}
index a i == Nothing
\end{code}
%
\noindent
then for all $0 \leq j < i$,
%
\begin{code}
index a j /= Nothing
\end{code}
\noindent
We use this invariant as a termination case in recursive algorithms over the
sub-values of a value.  (Rather than indexes into a data-structure, an
alternative representation is to use a zipper data structure~\cite{zipper} to
traverse data.  We have chosen explicit indexes to write simple tail-recursive
algorithms that can easily be transcribed to imperative languages.)

@@

Figure~\ref{fig:reduction} shows the reduction algorithm.  In this algorithm and
subsequent algorithms in the paper, functions in {\bf bold font} are left
undefined but their implementation is described in the text.  The function
\ttp{reduce} takes flags to customize the algorithm's behavior, a counterexample
\ttp{cex}, and the property \ttp{prop}.  The reduction begins at the first
proper sub-value of \ttp{cex}; call it \ttp{v}.  When the index \ttp{idx}
becomes out-of-bounds and returns \ttp{Nothing}, the algorithm terminates.
Otherwise, a list of new random values are generated.
%
\begin{code}
sizedArbitrary :: SubTypes a => Size -> a -> IO a
\end{code}
%
\noindent
generates a new value \ttp{v'} having the same type as \ttp{v} and that is
strictly smaller (with respect to the \ttp{size} method) than \ttp{v}.  Just
like QuickCheck's \ttp{arbitrary} method, \ttp{sizedArbitrary} generates
successively larger counterexamples when generating new values with which to
replace a sub-value.

@@

\subsubsection{Variant Counterexample Hypothesis}
A motivation for the design of the reduction algorithm is something we call the
\emph{variant counterexample hypothesis}: in the search space of possible values
from a given type \ttp{T}, if a known counterexample \ttp{cex} is a variant
\ttp{v} of \ttp{T}, then it is most probable that other counterexamples are also
from variant \ttp{v}.  As an example supporting the hypothesis, consider a
property about unintended variable capture over a language's parse tree
represented by a sum type with constructors for module imports, function
definitions, and global-variable assignments, respectively.  A function
definition counterexample can only be reduced to smaller function definition
counterexamples, the only construct in which variable capture is possible.

Recall that the algorithm begins at the \emph{first} sub-value of the
counterexample rather than the 0th sub-value.  The 0th sub-value of \ttp{cex} is
\ttp{cex} itself.  No invariant of the algorithm would be violated by beginning
with the 0th sub-value, and in particular, the algorithm would still terminate.
However, we begin with a strict sub-value because of the hypothesis.

The hypothesis may be incorrect for some properties, in which case SmartCheck
may potentially fail to discover a smaller counterexample.  However, in
Sections~\ref{sec:existential} and~\ref{sec:precondition}, we describe an
approaches to generalize counterexamples based on discovering new counterexample
variants.  These generalization techniques are executed in an (optional)
generalization phase, run after the reduction phase, in which this hypothesis is
implemented.

@@

We generalize the idea of ignoring primitive types by introducing the concept of
\emph{opaque types}.  If the reduction algorithm encounters a opaque type, it is
ignored.  Opaque types include the primitive types mentioned above, but in the
user can declare any substructure in a data type to be a opaque type by
providing custom instances.  Doing so effectively treats values from that type
as ``black boxes'', making SmartCheck more efficient if the user knows that some
portion of the structure cannot be shrunk or is irrelevant to the property.

Opaque types can be conditional.  For example, the user may want lists to be
shrunk in general, unless the element of the list are opaque themselves.  Such a
definition is possible.

@@

eval :: Exp -> Maybe Int
eval (C i) = Just i
eval (Add e0 e1) =
  liftM2 (+) (eval e0) (eval e1)
eval (Div e0 e1) =
  let e = eval e1 in
  if e == Just 0 then Nothing
    else liftM2 div (eval e0) e
\end{code}
%
\noindent
Now consider the property \ttp{prop\_div}, claiming that if \ttp{divSubTerms}
holds on an expression, then the evaluator returns \ttp{Just} a value:
%
\begin{code}
divSubTerms :: Exp -> Bool
divSubTerms (C _)         = True
odivSubTerms (Div _ (C 0)) = False
divSubTerms (Add e0 e1)   =  divSubTerms e0
                          && divSubTerms e1
divSubTerms (Div e0 e1)   =  divSubTerms e0
                          && divSubTerms e1

@@

Small counterexamples make debugging easier, but they are just half the battle.
To go from a specific counterexample to the required fix in a program, the
programmer must have a flash of insight in which she generalizes the
counterexample to a set of counterexamples for which the program and property
fails.  The generalization step is an important yet under-appreciated step in
the debugging process.  A characterizing formula reduces the noise in favor of
the signal by abstracting away portions of large counterexample that are
irrelevant to why it violates the property.

@@

@@

The extrapolation algorithm is shown in Figure~\ref{fig:universal}; let sketch
its specification.  The algorithm is similar to the reduction algorithm in
Figure~\ref{fig:reduction} (and in the implementation, the algorithms are
generalized and combined).  The function \ttp{extrapolate} returns a list of
indexes to be generalized in the original counterexample.  In the recursive
function \ttp{extrapolate'}, there is a function guard with a call
%
\begin{code}
subTree cex idx0 idx1
\end{code}
%
\noindent
where \ttp{subTree} has the type
%
\begin{code}
subTree :: SubTypes a => a -> Idx -> Idx -> Bool
\end{code}
%
\noindent
The value
%
\begin{code}
subTree cex idx0 idx1
\end{code}
%
is true if in \ttp{cex}, the value at index \ttp{idx0} is a child of index
\ttp{idx1} in a tree representation of \ttp{cex} (i.e., \ttp{subVals cex}).  The
\ttp{subTrees} guard prevents the algorithm from trying to generalize sub-values
that are abstracted away already since their parents have been generalized.
New sub-values are generated by \ttp{newVals}, shown in
Figure~\ref{fig:reduction}.

@@

We therefore introduce another kind of generalization we call \emph{existential
  sub-value generalization}.  In this generalization, if there is a
counterexample containing every possible variant as a sub-value, then we
abstract it.  For example, suppose that \ttp{divSubTerms} had no equation
%
\begin{code}
divSubTerms (Div _ (C 0)) = False
\end{code}
%
\noindent
Then the following would all counterexamples:
%
\begin{code}
Div (Add (C 3) (C 2)) (C 0)
Div (C 7) (Add (C (-5)) (C 5))
Div (C (-2)) (Div (C 0) (C 1))
\end{code}
%
\noindent
Because there are only three variants in the type, we have found counterexamples
built from each of them in the divisor.  We therefore can claim the following
formula holds:
%
\begin{code}
forall values x .
  forall constructors c .
    there exist arguments \(\stackrel{\rightarrow}{y}\) .
      such that Div x (c\(\stackrel{\rightarrow}{y}\))
\end{code}
%
\noindent
We therefore present an existential sub-value generalization algorithm that
performs constructor generalization.  Like with the other algorithms, this
algorithm also performs a breadth-first search over a counterexample.

@@

@@

\noindent
Recall that through the entire process, we never had to learn how GenI works,
what the property meant, or how to write a custom shrink method!

``Retro-fitting'' SmartCheck to GenI, which is a large software system, required
some boilerplate work.  Fortunately, because the program already uses
QuickCheck, \ttp{arbitrary} instances were already defined, but we had to hunt
down the data types used to write \ttp{instance Subtypes D}, for each data type
\ttp{D} (indeed, the job could have largely been done with Sed and Awk).

@@

\noindent
All benchmarks can be found
online.\footnote{\url{https://github.com/leepike/SmartCheck/tree/master/regression}}
We compare the size of the final counterexample returned (by counting
constructors) and the time required for counterexample generation and shrinking
in seconds.  The results are presented in Table~\ref{table:results}.  Again, we
summarize the mean, median, standard deviation, and the results at the 95th
percentile.  (While we provide the standard deviations, note that the plots are
not necessarily Gaussian.)

\begin{table}[ht]
\footnotesize
  \begin{center}
    \begin{tabular}{|r|r||c|c|}
\hline
\multicolumn{2}{|c||}{} & QuickCheck & SmartCheck \\
\hline \hline

\multirow{4}{*}{Reverse}
  & Mean       & 2, 0.002s  & 2, \sci{4.8}{4}s  \\
  \cline{2-4}
  & Std. dev.  & 0, 0.002   & 0, \sci{8.1}{4}  \\
  \cline{2-4}
  & 95\%       & 2, 0.007s  & 2, \sci{5.8}{4}s  \\
  \hline

\multirow{4}{*}{Div0}
  & Mean       & 10, 0.001s  & 5, 0.001s  \\
  \cline{2-4}
  & Std. dev.  & 6, 0.002    & 0, 0.001   \\
  \cline{2-4}
  & 95\%       & 23, 0.002s  & 5, 0.002s \\
  \hline

\multirow{4}{*}{Heap}
  & Mean       & 15, 0.016s  & 7, 0.007s  \\
  \cline{2-4}
  & Std. dev.  & 1, 0.025    & 2, 0.003   \\
  \cline{2-4}
  & 95\%       & 19, 0.045s  & 12, 0.015s  \\
  \hline

\multirow{4}{*}{Parser}
  & Mean       & 191, 0.013s  & 16, 0.241s  \\
  \cline{2-4}
  & Std. dev.  & 195, 0.018   & 75, 0.800   \\
  \cline{2-4}
  & 95\%       & 577, 0.044s  & 14, 0.521s  \\
  \hline

@@

  \end{center}
  \caption{Summarizing data for the graphs in Figure~\ref{fig:graphs}. Entries
    contain execution time (in seconds) and counterexample sizes (counting
    constructors).}
  \label{table:results}
\end{table}

>>>>

% can be 9pt
\documentclass{sigplanconf}
\usepackage{url}
\usepackage{tikz}
\usepackage{graphicx}
\usepackage{framed}
\usepackage{color}
\usepackage{url}
\usepackage{alltt}
\usepackage{tabularx}
\usepackage{subfigure}
\usepackage{fancyhdr}
\usepackage{paralist}
\usepackage{multirow}
\usepackage{pgf}
\usepackage{nth}

@@

%% \newcommand{\sci}[2]{\ensuremath{#1 \times 10^{#2}}}
\newcommand{\sci}[2]{\ensuremath{#1e^{#2}}}
\begin{document}

@@

That is, automated testing helps find a counterexample, but it may be a very
large counterexample.  We are left facing the \emph{shrinking problem}: assume
there exist a set of inputs $i_0, \ i_1, \ldots, \ i_n$, such that for program
$P$ (modeled as a function mapping inputs to output) and property $\phi$,
$\phi(P(i_0, \ i_1, \ldots, \ i_n))$ is false.  The shrinking problem is to find
the smallest set of inputs $i'_0, \ i'_1, \ldots, \ i'_n$, according to some
measure, such that $\phi(P(i\_0, \ i\_1, \ldots, \ i\_n))$ is also false.

@@

The third algorithm automatically strengthens properties by omitting ``similar''
counterexamples to the ones previously observed.  The algorithm is motivated by
noting that there are often multiple ways in which a property may fail; for
example, a property stating that pretty-printing an AST and then parsing it
results in the original AST may fail due to multiple bugs, such that each bug in
isolation is sufficient to cause failure.  During testing, it is useful to
discover counterexamples arising from all the bugs in one go.  In practice, the
problem is solved by discovering a counterexample \ttp{cex}, abstracting it, and
then adding a new precondition to the property that informally says ``omit
counterexamples of form \ttp{cex}.''  Adding preconditions manually is laborious
and may cause the programmer to make premature fixes to the program, if she
believes she has isolated the error before she actually does.

@@

The most recent released version of QuickCheck provides a \ttp{genericShrink}
function for algebraic data types, so the user merely has to declare
%
\begin{code}
shrink = genericShrink
\end{code}
%
\noindent
to obtain shrink instances.  (The implementation relies on the same generics
package that SmartCheck does; see Section~\ref{sec:preliminaries}.)

@@

\begin{table}[ht]
\footnotesize
  \begin{center}
    \begin{tabular}{|r||c|c|c|c|c|c|}
\hline

              & \multicolumn{2}{c}{Mean} & \multicolumn{2}{c}{Std. dev.} &
\multicolumn{2}{c}{95\%}\\\hline
              & size & time  & size & time & size & time \\\hline\hline
QC none       & 83 & 0.013   & 22 & 0.014   & 119 & 0.023\\\hline
QC trunc      & 40 & 0.064   & 20 & 0.334   & 84 & 0.153\\\hline
QC tuple      & 10 & 0.031   & 4 &  0.039   & 16 & 0.052\\\hline
QC generic    & &        &        &           & &\\\hline
SmartCheck    & 6 & 0.021s    &  4 & 0.009   & 13 & 0.036\\\hline

    \end{tabular}
  \end{center}
  \caption{Summarizing data for the graph in Figure~\ref{fig:graphs}. Entries
    contain execution time (in seconds) and counterexample sizes (counting
    \ttp{Int16} values).}
  \label{table:results}
\end{table}

The results of our discussion above are summarized in Table~\ref{table:results}
and the corresponding graph in Figure~\ref{fig:graphs}. They show four
approaches to shrinking with QuickCheck: no shrinking (QC none), to provide a
baseline, truncated shrinking (QC trunc) with \ttp{Int16} shrinking disabled, in
which each list is truncated to 10 elements, tuple shrinking (QC tuple), also
with \ttp{Int16} shrinking disabled, and finally shrinking using QuickCheck's
\ttp{genericShrink} (qc generic).  We omit SmallCheck, since as noted, it is not
even feasible for the example with default instances.\footnote{All results
  reported in the paper are with a GHC-7.6.3-compiled program, using \ttp{-O2},
  on a 2.8 GHz 4~core i7 with 16GB memory under OSX v. 10.9.2.}  We graph the
final counterexample size, measured in the total number of \ttp{Int16} values,
and execution time for counterexample discovery and shrinking.
Figure~\ref{fig:graphs} shows how many of the 1000 runs resulted in a final
value of a particular size, fitting the data with a Bezier curve.

@@

class Arbitrary a => SubTypes a where
  size    :: a -> Size
  index   :: a -> Idx -> Maybe SubVal
  replace :: a -> Idx -> SubVal -> a
  constr  :: a -> String
  constrs :: a -> [String]
  opaque  :: a -> Bool
  subVals :: a -> Tree SubVal
\end{code}
%
\noindent
The \ttp{SubTypes} type class requires QuickCheck's \ttp{Arbitrary} as a
super-class.  \ttp{SubTypes} has the following methods:
\begin{itemize}
\item \ttp{size} returns the size of a value---intuitively, the number of
  constructors contained within it,
\item \ttp{index} returns a \emph{sub-value} at a breadth-first index in a
  value,
\item \ttp{replace} replaces a sub-value at a particular focus (returning the
  original value if the index is out-of-bounds),
\item \ttp{constr} returns a string representation of the constructor tagging
  the value,
\item \ttp{constrs} returns the list of all possible constructors from the
  value's type,
\item \ttp{opaque} is false when the type of the value is an ``interesting
  type''; informally, this is a type other than a primitive type like \ttp{Int},
  \ttp{Char}, or \ttp{Bool}, and may be user-defined.  See
  Section~\ref{sec:base} for a full discussion.
\item \ttp{subVals} returns a tree of all non opaque-type sub-values.  A tree
  has the type
%
  \begin{code}
data Tree a = Node \{ rootLabel :: a
                   , subForest :: [Tree a] \}
  \end{code}
%
\end{itemize}

@@

subVals (B (B L\sub{0} L\sub{1}) L\sub{2}) =
  Node (B (B L\sub{0} L\sub{1}) L\sub{2})
       [ Node (B L\sub{0} L\sub{1}) [Node L\sub{0} [], Node L\sub{1} []]
       , Node L\sub{2} []
       ]
\end{code}
%
\noindent
The \ttp{SubVal} type is an existential data type, used as a generic container
for sub-values from a counterexample.  We will sometimes refer to the unwrapped
value returned by \ttp{index a i} as the \emph{$i$th sub-value of $a$}, so for
example, (\ttp{B\sub{4} L\sub{6} L\sub{8}}) is the 4th sub-value of \ttp{tree}.
An invariant of \ttp{index} is that for any value \ttp{a}, and for the smallest
$i \geq 0$ such that
%
\begin{code}
index a i == Nothing
\end{code}
%
\noindent
then for all $0 \leq j < i$,
%
\begin{code}
index a j /= Nothing
\end{code}
\noindent
We use this invariant as a termination case in recursive algorithms over the
sub-values of a value.  (Rather than indexes into a data-structure, an
alternative representation is to use a zipper data structure~\cite{zipper} to
traverse data.  We have chosen explicit indexes to write simple tail-recursive
algorithms that can easily be transcribed to imperative languages.)

@@

Figure~\ref{fig:reduction} shows the reduction algorithm.  In this algorithm and
subsequent algorithms in the paper, functions in {\bf bold font} are left
undefined but their implementation is described in the text.  The function
\ttp{reduce} takes flags to customize the algorithm's behavior, a counterexample
\ttp{cex}, and the property \ttp{prop}.  The reduction begins at the first
proper sub-value of \ttp{cex}; call it \ttp{v} (this is an optimization
described in Section~\ref{sec:variant}).  When the index \ttp{idx} becomes
out-of-bounds and returns \ttp{Nothing}, the algorithm terminates.  Otherwise, a
list of new random values are generated.
%
\begin{code}
sizedArbitrary :: SubTypes a => Size -> a -> IO a
\end{code}
%
\noindent
generates a new value \ttp{v'} having the same type as \ttp{v} and that is
strictly smaller (with respect to the \ttp{size} method) than \ttp{v}.  Just
like QuickCheck's \ttp{arbitrary} method, \ttp{sizedArbitrary} generates
successively larger counterexamples when generating new values with which to
replace a sub-value.

@@

\subsubsection{Variant Counterexample Hypothesis}\label{sec:variant}
A motivation for the design of the reduction algorithm is something we call the
\emph{variant counterexample hypothesis}: in the search space of possible values
from a given type \ttp{T}, if a known counterexample \ttp{cex} is a variant
\ttp{v} of \ttp{T}, then it is most probable that other counterexamples are also
from variant \ttp{v}.  As an example supporting the hypothesis, consider a
property about unintended variable capture over a language's parse tree
represented by a sum type with constructors for module imports, function
definitions, and global-variable assignments, respectively.  A function
definition counterexample can only be reduced to smaller function definition
counterexamples, the only construct in which variable capture is possible.

Recall that the algorithm begins at the \emph{first} sub-value of the
counterexample rather than the zeroth sub-value.  The zeroth sub-value of
\ttp{cex} is \ttp{cex} itself.  No invariant of the algorithm would be violated
by beginning with the zeroth sub-value, and in particular, the algorithm would
still terminate.  However, we begin with a strict sub-value because of the
hypothesis.

The hypothesis may be incorrect for some properties, in which case SmartCheck
may potentially fail to discover a smaller counterexample.  However, in
Sections~\ref{sec:existential} and~\ref{sec:precondition}, we describe 
approaches to generalize counterexamples based on discovering new counterexample
variants.  These generalization techniques are executed in an (optional)
generalization phase, run after the reduction phase, in which this hypothesis is
implemented.

@@

We generalize the idea of ignoring primitive types by introducing the concept of
\emph{opaque types}.  If the reduction algorithm encounters an opaque type, it is
ignored.  Opaque types include the primitive types mentioned above, but in the
user can declare any substructure in a data type to be an opaque type by
providing custom instances.  Doing so effectively treats values from that type
as ``black boxes'', making SmartCheck more efficient if the user knows that some
portion of the structure cannot be shrunk or is irrelevant to the property.

Opaque types can be conditional.  For example, the user may want lists to be
shrunk in general, unless the elements of the list are opaque themselves.  Such a
definition is possible.

@@

eval :: Exp -> Maybe Int
eval (C i) = Just i
eval (Add e0 e1) =
  liftM2 (+) (eval e0) (eval e1)
eval (Div e0 e1) =
  let e = eval e1 in
  if e == Just 0 then Nothing
    else liftM2 div (eval e0) e
\end{code}
%
\noindent
Now consider the property \ttp{prop\_div}, claiming that if \ttp{divSubTerms}
holds on an expression, then the evaluator returns \ttp{Just} a value:
%
\begin{code}
divSubTerms :: Exp -> Bool
divSubTerms (C _)         = True
divSubTerms (Div _ (C 0)) = False
divSubTerms (Add e0 e1)   =  divSubTerms e0
                          && divSubTerms e1
divSubTerms (Div e0 e1)   =  divSubTerms e0
                          && divSubTerms e1

@@

Small counterexamples make debugging easier, but they are just half the battle.
To go from a specific counterexample to the required fix in a program, the
programmer must have a flash of insight in which she generalizes the
counterexample to a set of counterexamples for which the program and property
fails.  The generalization step is an important yet under-appreciated step in
the debugging process.  A characterizing formula reduces the noise in favor of
the signal by abstracting away portions of a large counterexample that are
irrelevant to why it violates the property.

@@

In SmartCheck, the universal and existential generalization phases run after a
counterexample has been minimized.  Precondition strengthening is used when
iteratively generating multiple counterexamples, so it is interspersed with
counterexample reduction.

@@

The extrapolation algorithm is shown in Figure~\ref{fig:universal}; let us sketch
its specification.  The algorithm is similar to the reduction algorithm in
Figure~\ref{fig:reduction} (and in the implementation, the algorithms are
generalized and combined).  The function \ttp{extrapolate} returns a list of
indexes to be generalized in the original counterexample.  In the recursive
function \ttp{extrapolate'}, there is a function guard with a call
%
\begin{code}
subTree cex idx0 idx1
\end{code}
%
\noindent
where \ttp{subTree} has the type
%
\begin{code}
subTree :: SubTypes a => a -> Idx -> Idx -> Bool
\end{code}
%
\noindent
The value
%
\begin{code}
subTree cex idx0 idx1
\end{code}
%
is true if in \ttp{cex}, the value at index \ttp{idx0} is a child of index
\ttp{idx1} in a tree representation of \ttp{cex} (i.e., \ttp{subVals cex}).  The
\ttp{subTrees} guard prevents the algorithm from trying to generalize sub-values
that are abstracted away already since their parents have been generalized.
New sub-values are generated by \ttp{newVals}, shown in
Figure~\ref{fig:reduction}.

@@

We therefore introduce another kind of generalization we call \emph{existential
  sub-value generalization}.  In this generalization, if there is a
counterexample containing every possible variant as a sub-value, then we
abstract it.  For example, suppose that \ttp{divSubTerms} had no equation
%
\begin{code}
divSubTerms (Div _ (C 0)) = False
\end{code}
%
\noindent
Then the following would all be counterexamples:
%
\begin{code}
Div (Add (C 3) (C 2)) (C 0)
Div (C 7) (Add (C (-5)) (C 5))
Div (C (-2)) (Div (C 0) (C 1))
\end{code}
%
\noindent
Because there are only three variants in the type, we have found counterexamples
built from each of them in the divisor.  We therefore can claim the following
formula holds:
%
\begin{code}
forall values x .
  forall constructors c .
    there exist arguments \(\stackrel{\rightarrow}{y}\) .
      such that Div x (c\(\stackrel{\rightarrow}{y}\))
\end{code}
%
\noindent
We therefore present an existential sub-value generalization algorithm that
performs constructor generalization.  Like with the other algorithms, this
algorithm also performs a breadth-first search over a counterexample.

@@

SmartCheck can run in a real-eval-print loop (REPL), discovering new
counterexamples, shrinking them, generalizing them, and then repeating.  In each
iteration of the REPL, the property precondition is strengthened, requiring that
\ttp{matchesShape} does not hold on the current counterexample candidate and
\emph{any} previous counterexample generalization already discovered.

@@

\noindent
Through the entire process, we never had to learn how GenI works, what the
property meant, or how to write a custom shrink method!

%% ``Retro-fitting'' SmartCheck to GenI, which is a large software system, required
%% some boilerplate work.  Fortunately, because the program already uses
%% QuickCheck, \ttp{arbitrary} instances were already defined, but we had to hunt
%% down the data types used to write \ttp{instance Subtypes D}, for each data type
%% \ttp{D} (indeed, the job could have largely been done with Sed and Awk).

@@

\noindent
All benchmarks can be found
online.\footnote{\url{https://github.com/leepike/SmartCheck/tree/master/regression}}
We compare the size of the final counterexample returned (by counting
constructors) and the time required for counterexample generation and shrinking
in seconds.  The results are presented in Table~\ref{table:benchmarks}.  Again, we
summarize the mean, median, standard deviation, and the results at the 95th
percentile.  (While we provide the standard deviations, note that the plots are
not necessarily Gaussian.)

\todo{generic or hand-written?}

\begin{table}[ht]
\scriptsize
  \begin{center}
    \begin{tabular}{|r|r||c|c|c|c|c|c|}

\hline
  \multicolumn{2}{|c||}{} & \multicolumn{2}{c|}{Mean} & \multicolumn{2}{c|}{Std. dev.} &
\multicolumn{2}{c|}{95\%}\\\cline{3-8}
  \multicolumn{2}{|c||}{} & size & time & size & time & size & time \\\hline\hline

\multirow{2}{*}{Reverse}
  & QC & 2 & 0.002         & 0 & 0.002         & 2 & 0.007 \\\cline{2-8}
  & SC & 2 & \sci{4.8}{-4} & 0 & \sci{8.1}{-4} & 2 & \sci{5.8}{-4}\\\hline\hline

\multirow{2}{*}{Div0}
  & QC & 10 & 0.001 & 6 & 0.002 & 23 & 0.002 \\\cline{2-8}
  & SC & 5 & 0.001  & 0 & 0.001 & 5 & 0.002  \\\hline\hline

\multirow{2}{*}{Heap}
  & QC & 15 & 0.016  & 1 & 0.025 & 19 & 0.045 \\\cline{2-8}
  & SC & 7  & 0.007  & 2 & 0.003 & 12 & 0.015 \\\hline\hline

\multirow{2}{*}{Parser}
  & QC &  191 & 0.013 & 195 & 0.018 & 577 & 0.044 \\\cline{2-8}
  & SC &  16  & 0.241 & 75  & 0.800 & 14  & 0.521 \\\hline

@@

  \end{center}
  \caption{Summarizing data for the graphs in Figure~\ref{fig:graphs}. Entries
    contain execution time (in seconds) and counterexample sizes (counting
    constructors).}
  \label{table:benchmarks}
\end{table}
